# Exploratory Data Analysis

::: {.callout-note title="Connection with the theoretical volume"}
Probability, statistics, covariance, and multivariate representation are developed in greater depth in Chapters 2 and 3 of *Mathematical Foundations of Machine Learning* [@rodriguez2026fundamentos].
:::

## Chapter objectives

By the end of this chapter, the reader will be able to inspect a prepared dataset, compute frequencies, build exploratory plots, and identify patterns that may affect later modeling decisions.

## ATUS exploratory analysis

```{r}
library(readr)
library(dplyr)
library(ggplot2)

atus_ml <- read_csv("../datos/atus_ml_preparado.csv", show_col_types = FALSE)

atus_ml <- atus_ml |>
  mutate(
    ID_HORA_NUM = as.numeric(ID_HORA),
    MES_NUM = as.numeric(MES),
    accidente_con_victimas = factor(
      accidente_con_victimas,
      levels = c("Con víctimas", "Solo daños")
    )
  )
```

### Outcome distribution

```{r}
tabla_clase <- table(atus_ml$accidente_con_victimas)
round(100 * prop.table(tabla_clase), 2)
```

::: {.callout-tip title="Interpretation"}
Class imbalance matters. A model can achieve a deceptively high overall accuracy while performing poorly on the less frequent class.
:::

### Accidents by hour

```{r fig.width=7, fig.height=4.5}
atus_ml |>
  count(ID_HORA_NUM, name = "n") |>
  ggplot(aes(x = ID_HORA_NUM, y = n)) +
  geom_col() +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "Number of accidents by hour of day",
    x = "Hour",
    y = "Number of accidents"
  ) +
  theme_minimal()
```

## COVID-19 exploratory analysis

The prepared COVID-19 file is detected dynamically so that the chapter does not depend on a single hard-coded year.

```{r}
archivos_covid <- list.files(
  "../datos/covid19/procesados",
  pattern = "covid19_mexico_[0-9]{4}_ml_preparado\\.csv\\.gz$",
  full.names = TRUE
)

if (length(archivos_covid) == 0) {
  stop("No prepared COVID-19 dataset was found.")
}

covid <- read_csv(archivos_covid[1], show_col_types = FALSE)
```

### Age distribution

```{r fig.width=7, fig.height=4.8}
ggplot(covid, aes(x = EDAD)) +
  geom_histogram(bins = 35) +
  labs(
    title = "Age distribution in the educational COVID-19 dataset",
    x = "Age",
    y = "Frequency"
  ) +
  theme_minimal()
```

### Comorbidities

```{r}
covid |>
  count(NUM_COMORBILIDADES) |>
  arrange(NUM_COMORBILIDADES)
```

::: {.callout-important title="Do not confuse sample proportions with epidemiological rates"}
If the educational sample was balanced or stratified, observed class proportions in this chapter should not be interpreted as population mortality or prevalence estimates.
:::

## Chapter summary

Exploratory analysis helps reveal imbalance, unusual values, concentration patterns, and possible data-quality issues before fitting models. Its role is descriptive and diagnostic; it does not, by itself, establish causal relationships.
