# Logistic Regression for Classification

::: {.callout-note title="Connection with the theoretical volume"}
Logistic regression, maximum likelihood, and probabilistic classification are developed in greater depth in Chapters 3, 6, and 11 of *Mathematical Foundations of Machine Learning* [@rodriguez2026fundamentos].
:::

## Chapter objectives

By the end of this chapter, the reader will be able to fit a logistic-regression model, interpret probabilities and coefficients, choose a classification threshold, and distinguish discrimination from calibration.

## Mathematical foundation

Logistic regression models a probability through

$$
p(x)=\frac{1}{1+e^{-z}},
$$

where

$$
z=\beta_0+\beta_1x_1+\cdots+\beta_px_p.
$$

Equivalently,

$$
\log\left(\frac{p}{1-p}\right)=\beta_0+\beta_1x_1+\cdots+\beta_px_p.
$$

The expression on the left is the *logit*.

## Logistic curve

```{r fig.width=7, fig.height=4.5}
library(ggplot2)

z <- seq(-10, 10, length.out = 200)
logistic_data <- data.frame(
  z = z,
  p = 1 / (1 + exp(-z))
)

ggplot(logistic_data, aes(x = z, y = p)) +
  geom_line(linewidth = 1.2) +
  labs(
    title = "Logistic function",
    subtitle = "Maps real-valued scores to probabilities",
    x = "z",
    y = "Probability"
  ) +
  theme_minimal()
```

## ATUS case: accidents with victims

```{r}
library(readr)
library(dplyr)

atus_ml <- read_csv("../datos/atus_ml_preparado.csv", show_col_types = FALSE)

atus_model <- atus_ml |>
  mutate(
    victims_binary = ifelse(accidente_con_victimas == "Con víctimas", 1, 0),
    MES = as.factor(MES),
    ID_HORA = as.numeric(ID_HORA),
    DIASEMANA = as.factor(DIASEMANA),
    TIPACCID = as.factor(TIPACCID),
    CAUSAACCI = as.factor(CAUSAACCI)
  )

set.seed(2026)
idx <- sample(seq_len(nrow(atus_model)), size = floor(0.70 * nrow(atus_model)))
train <- atus_model[idx, ]
test <- atus_model[-idx, ]

model_logit <- glm(
  victims_binary ~ MES + ID_HORA + DIASEMANA + TIPACCID + CAUSAACCI,
  data = train,
  family = binomial
)
```

::: {.callout-tip title="Interpreting coefficients"}
A coefficient describes a change in log-odds while holding the remaining predictors constant. Its exponential, $e^{\beta_j}$, can be interpreted as an odds ratio.
:::

## Probabilities and classification threshold

```{r}
test$prob_victims <- predict(model_logit, newdata = test, type = "response")
threshold <- 0.30

test$predicted_class <- ifelse(
  test$prob_victims >= threshold,
  "Con víctimas",
  "Solo daños"
)
```

::: {.callout-note title="The threshold is not part of the fitted probability model"}
The logistic model produces probabilities. A threshold is a later decision rule and changes the sensitivity-specificity trade-off.
:::

## Discrimination and calibration

A useful probabilistic classifier must distinguish two ideas:

- **discrimination:** how well the model ranks lower-risk and higher-risk observations;
- **calibration:** how closely predicted probabilities agree with observed frequencies.

A model can discriminate well while still producing poorly calibrated probabilities.

## COVID-19 educational case

The COVID-19 route dynamically uses the processed file selected by the reproducible preparation procedure instead of fixing a specific year.

```{r}
covid_files <- list.files(
  "../datos/covid19/procesados",
  pattern = "covid19_mexico_[0-9]{4}_ml_preparado\\.csv\\.gz$",
  full.names = TRUE
)

if (length(covid_files) == 0) {
  stop("No prepared COVID-19 dataset was found.")
}

covid <- read_csv(covid_files[1], show_col_types = FALSE)

covid_model <- covid |>
  select(
    MURIO,
    EDAD,
    NEUMONIA,
    DIABETES,
    HIPERTENSION,
    OBESIDAD,
    RENAL_CRONICA
  ) |>
  tidyr::drop_na()
```

::: {.callout-important title="Predicted probability is not automatically population risk"}
If the educational dataset over-samples deaths or otherwise alters the class proportion, raw logistic probabilities do not directly represent population prevalence. Balanced samples may still be useful for teaching classification, but a risk simulator requires natural-prevalence data or appropriate weighting and calibration.
:::

```{r}
set.seed(2026)
idx_covid <- sample(
  seq_len(nrow(covid_model)),
  size = floor(0.80 * nrow(covid_model))
)

covid_train <- covid_model[idx_covid, ]
covid_test <- covid_model[-idx_covid, ]

covid_logit <- glm(
  MURIO ~ EDAD + NEUMONIA + DIABETES + HIPERTENSION + OBESIDAD + RENAL_CRONICA,
  data = covid_train,
  family = binomial
)

covid_test$sample_probability <- predict(
  covid_logit,
  newdata = covid_test,
  type = "response"
)
```

::: {.callout-warning title="Educational use only"}
This model is not a diagnostic or prognostic tool. Its coefficients describe associations within the dataset and depend on time period, coverage, data quality, and sampling strategy.
:::

## Chapter summary

Logistic regression separates two stages: probability estimation and, when needed, class assignment through a threshold. Responsible probability interpretation requires attention to prevalence, sampling, and calibration, especially in health applications.
