# Naive Bayes Classifier

::: {.callout-note title="Connection with the theoretical volume"}
The mathematical treatment of conditional probability, Bayes' theorem, and probabilistic classification is developed more deeply in Chapters 3 and 6 of *Mathematical Foundations of Machine Learning* [@rodriguez2026fundamentos].
:::

## Learning objectives

By the end of this chapter, the reader will be able to:

- explain Bayes' theorem in a classification setting;
- interpret priors, likelihoods, and posterior probabilities;
- describe the conditional-independence assumption;
- distinguish categorical and Gaussian Naive Bayes;
- understand the zero-probability problem and Laplace smoothing;
- fit and evaluate a Naive Bayes classifier in R;
- select a smoothing value using validation rather than the final test set;
- interpret posterior scores cautiously when prevalence has been modified.

## Bayes' theorem

For a class $C$ and observed predictors $X$,

$$
P(C\mid X)=\frac{P(X\mid C)P(C)}{P(X)}.
$$

For classification, $P(X)$ is common to all classes, so we compare

$$
P(C\mid X)\propto P(X\mid C)P(C).
$$

The components are:

- $P(C)$: prior probability of the class;
- $P(X\mid C)$: likelihood of the predictors under the class;
- $P(C\mid X)$: posterior probability after observing the predictors.

## The naive conditional-independence assumption

For predictors $X_1,\ldots,X_p$, Naive Bayes assumes conditional independence given the class:

$$
P(X_1,\ldots,X_p\mid C)=\prod_{j=1}^{p}P(X_j\mid C).
$$

Therefore,

$$
P(C\mid X_1,\ldots,X_p)
\propto P(C)\prod_{j=1}^{p}P(X_j\mid C).
$$

::: {.callout-tip title="Key idea"}
Naive Bayes constructs one score for each class from the prior and the class-conditional predictor distributions. The largest score determines the predicted class.
:::

## Manual categorical example

```{r nb-manual-data}
weather <- c("Dry", "Dry", "Rain", "Rain", "Dry",
             "Rain", "Dry", "Rain", "Dry", "Rain")
time <- c("Day", "Night", "Day", "Night", "Day",
          "Night", "Night", "Day", "Day", "Night")
victims <- factor(c("No", "No", "Yes", "Yes", "No",
                    "Yes", "Yes", "No", "No", "Yes"))

nb_demo <- data.frame(weather, time, victims)

priors <- prop.table(table(nb_demo$victims))
p_weather <- prop.table(table(nb_demo$weather, nb_demo$victims), margin = 2)
p_time <- prop.table(table(nb_demo$time, nb_demo$victims), margin = 2)

score_yes <- priors["Yes"] * p_weather["Rain", "Yes"] * p_time["Night", "Yes"]
score_no <- priors["No"] * p_weather["Rain", "No"] * p_time["Night", "No"]

posterior <- c(Yes = score_yes, No = score_no)
posterior / sum(posterior)
```

The raw products are proportional to the posterior probabilities. Dividing by their sum normalizes them to add to one.

## Zero probabilities and Laplace smoothing

If a categorical level never appears with a class in the training data, an estimated conditional probability may be zero. Multiplication then forces the entire class score to zero.

Laplace smoothing replaces raw frequencies with adjusted counts:

$$
\widehat{P}(X_j=x\mid C=c)=\frac{n_{x,c}+\alpha}{n_c+\alpha k},
$$

where $k$ is the number of categories. A common value is $\alpha=1$, but it should be treated as a modeling choice rather than a universal constant.

## Gaussian Naive Bayes

For a continuous predictor, a common assumption is

$$
X_j\mid C=c\sim N(\mu_{jc},\sigma_{jc}^{2}).
$$

The mean and variance are estimated separately for each predictor within each class.

## ATUS route

```{r nb-atus-load}
library(readr)
library(dplyr)
library(e1071)
library(ggplot2)
source("../util_graficas.R")

atus <- read_csv("../datos/atus_ml_preparado.csv", show_col_types = FALSE)

set.seed(2026)
atus_nb <- atus |>
  sample_n(min(20000, nrow(atus))) |>
  transmute(
    accident_class = factor(
      accidente_con_victimas,
      levels = c("Solo daños", "Con víctimas"),
      labels = c("Property damage only", "With casualties")
    ),
    MES = factor(MES),
    ID_HORA = as.numeric(ID_HORA),
    DIASEMANA = factor(DIASEMANA),
    TIPACCID = factor(TIPACCID),
    CAUSAACCI = factor(CAUSAACCI)
  ) |>
  na.omit()
```

## Stratified train, validation, and test split

The final test set must not be used to choose Laplace smoothing or other modeling options.

```{r nb-split}
stratified_sample <- function(y, proportion, seed) {
  set.seed(seed)
  unlist(lapply(
    split(seq_along(y), y),
    function(i) sample(i, size = floor(proportion * length(i)))
  ))
}

idx_trainval <- stratified_sample(atus_nb$accident_class, 0.80, 2026)
trainval <- atus_nb[idx_trainval, ]
test_nb <- atus_nb[-idx_trainval, ]

idx_train <- stratified_sample(trainval$accident_class, 0.75, 2027)
train_nb <- trainval[idx_train, ]
valid_nb <- trainval[-idx_train, ]

prop.table(table(train_nb$accident_class))
prop.table(table(valid_nb$accident_class))
prop.table(table(test_nb$accident_class))
```

## Common metric function

```{r nb-metrics-function}
binary_metrics_nb <- function(actual, predicted,
                              positive = "With casualties") {
  levels_all <- c("Property damage only", "With casualties")
  actual <- factor(actual, levels = levels_all)
  predicted <- factor(predicted, levels = levels_all)
  m <- table(Actual = actual, Predicted = predicted)

  tp <- m[positive, positive]
  fn <- m[positive, "Property damage only"]
  fp <- m["Property damage only", positive]
  tn <- m["Property damage only", "Property damage only"]

  safe_div <- function(a, b) if (b == 0) NA_real_ else as.numeric(a / b)

  sensitivity <- safe_div(tp, tp + fn)
  specificity <- safe_div(tn, tn + fp)
  precision <- safe_div(tp, tp + fp)
  accuracy <- safe_div(tp + tn, sum(m))
  f1 <- if (is.na(precision) || is.na(sensitivity) ||
            precision + sensitivity == 0) NA_real_ else
    2 * precision * sensitivity / (precision + sensitivity)

  data.frame(
    accuracy = accuracy,
    sensitivity = sensitivity,
    specificity = specificity,
    balanced_accuracy = mean(c(sensitivity, specificity), na.rm = TRUE),
    precision = precision,
    f1 = f1
  )
}
```

## Select Laplace smoothing on validation data

```{r nb-select-laplace}
laplace_grid <- c(0, 0.5, 1, 2)

validation_nb <- do.call(rbind, lapply(laplace_grid, function(alpha) {
  model <- e1071::naiveBayes(
    accident_class ~ MES + ID_HORA + DIASEMANA + TIPACCID + CAUSAACCI,
    data = train_nb,
    laplace = alpha
  )

  pred <- predict(model, valid_nb, type = "class")
  cbind(laplace = alpha, binary_metrics_nb(valid_nb$accident_class, pred))
}))

validation_nb
best_laplace <- validation_nb$laplace[
  which.max(validation_nb$balanced_accuracy)
]
best_laplace
```

::: {.callout-important title="Why validation matters"}
Trying several smoothing values on the final test set and then reporting the best one would make the test result optimistic. The test set is reserved for the model selected before seeing its performance.
:::

## Refit and evaluate once on the test set

```{r nb-final-model}
final_nb <- e1071::naiveBayes(
  accident_class ~ MES + ID_HORA + DIASEMANA + TIPACCID + CAUSAACCI,
  data = trainval,
  laplace = best_laplace
)

pred_test_nb <- predict(final_nb, test_nb, type = "class")
prob_test_nb <- predict(final_nb, test_nb, type = "raw")

confusion_nb <- table(
  Actual = test_nb$accident_class,
  Predicted = pred_test_nb
)

confusion_nb
binary_metrics_nb(test_nb$accident_class, pred_test_nb)
head(prob_test_nb)
```

## Inspect learned priors and conditional distributions

```{r nb-inspect}
final_nb$apriori
final_nb$tables$MES
```

::: {.callout-important title="Posterior scores are model-dependent"}
`type = "raw"` returns posterior scores under the fitted assumptions and training distribution. They are not automatically calibrated real-world probabilities. Calibration must be checked separately, especially when class prevalence has been modified.
:::

## Interactive laboratory: Naive Bayes posterior probability

::: {.content-visible when-format="html"}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 700
#| components: [viewer]

library(shiny)

ui <- fluidPage(
  titlePanel("Naive Bayes calculator"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("prior", "P(Class 1)", 0.05, 0.95, 0.40, step = 0.05),
      sliderInput("e1", "P(Evidence | Class 1)", 0.01, 0.99, 0.80, step = 0.01),
      sliderInput("e0", "P(Evidence | Class 0)", 0.01, 0.99, 0.25, step = 0.01),
      radioButtons("observed", "Observed evidence",
                   choices = c("Yes" = "yes", "No" = "no"))
    ),
    mainPanel(
      tableOutput("calculation"),
      plotOutput("plot", height = "340px"),
      textOutput("decision")
    )
  )
)

server <- function(input, output, session) {
  posterior <- reactive({
    p1 <- input$prior
    p0 <- 1 - p1
    if (input$observed == "yes") {
      l1 <- input$e1; l0 <- input$e0
    } else {
      l1 <- 1 - input$e1; l0 <- 1 - input$e0
    }
    n1 <- p1 * l1; n0 <- p0 * l0
    evidence <- n1 + n0
    c(Class_0 = n0 / evidence, Class_1 = n1 / evidence)
  })

  output$calculation <- renderTable({
    p <- posterior()
    data.frame(Class = names(p), Posterior = round(as.numeric(p), 4))
  })

  output$plot <- renderPlot({
    p <- posterior()
    barplot(p, ylim = c(0, 1), ylab = "Posterior probability",
            main = "Naive Bayes result")
    abline(h = 0.5, lty = 2)
  })

  output$decision <- renderText({
    p <- posterior()
    paste("Predicted class:",
          ifelse(p["Class_1"] >= p["Class_0"], "Class 1", "Class 0"))
  })
}

shinyApp(ui, server)
```
:::

::: {.content-visible when-format="pdf"}
The web edition includes an interactive calculator for prior, likelihood, and posterior probabilities.
:::

## COVID-19 educational route

The processed period is selected using the same reproducible comparison table used in the data-preparation chapter.

```{r nb-covid-load}
comparison_file <- "../datos/covid19/comparacion_estructural_2020_2025.csv"
comparison <- read_csv(comparison_file, show_col_types = FALSE)

selected_year <- comparison |>
  filter(is_usable_for_book, !is.na(prepared_ml_file)) |>
  arrange(desc(common_columns), desc(rows)) |>
  slice(1) |>
  pull(year)

covid_file <- file.path(
  "../datos/covid19/procesados",
  paste0("covid19_mexico_", selected_year, "_ml_preparado.csv.gz")
)

covid_nb <- read_csv(covid_file, show_col_types = FALSE) |>
  select(
    MURIO, EDAD, NEUMONIA, DIABETES, HIPERTENSION,
    OBESIDAD, RENAL_CRONICA, NUM_COMORBILIDADES
  ) |>
  tidyr::drop_na() |>
  mutate(
    outcome = factor(
      MURIO,
      levels = c(0, 1),
      labels = c("No recorded death", "Recorded death")
    ),
    NEUMONIA = factor(NEUMONIA),
    DIABETES = factor(DIABETES),
    HIPERTENSION = factor(HIPERTENSION),
    OBESIDAD = factor(OBESIDAD),
    RENAL_CRONICA = factor(RENAL_CRONICA)
  )
```

A Naive Bayes model can combine Gaussian treatment of `EDAD` and `NUM_COMORBILIDADES` with categorical treatment of the binary comorbidity variables.

::: {.callout-warning title="Educational use only"}
Posterior scores from this administrative educational sample must not be used for diagnosis, prognosis, or medical decisions. If the class distribution has been intentionally altered, the posterior probabilities do not directly represent population risk.
:::

## Supplementary materials for this chapter

| Resource | Purpose | Open or play | Download |
|---|---|---|---|
| Explanatory video | Audiovisual explanation of Naive Bayes. | [Watch on YouTube](https://www.youtube.com/watch?v=UbfJ3bLDwUI){target="_blank"} | — |
| PDF presentation | Slides for study or classroom presentation. | [View PDF](../recursos/capitulo-11/capitulo-11-naive-bayes-presentacion.pdf){target="_blank"} | [Download PDF](../recursos/capitulo-11/capitulo-11-naive-bayes-presentacion.pdf){download="chapter-11-naive-bayes-presentation.pdf"} |
| Editable presentation | PowerPoint file for classroom use or adaptation. | [Open PPTX](../recursos/capitulo-11/capitulo-11-naive-bayes-presentacion.pptx) | [Download PPTX](../recursos/capitulo-11/capitulo-11-naive-bayes-presentacion.pptx){download="chapter-11-naive-bayes-presentation.pptx"} |
| Infographic | Visual summary of the algorithm. | [View infographic](../recursos/capitulo-11/capitulo-11-naive-bayes-infografia.png){target="_blank"} | [Download PNG](../recursos/capitulo-11/capitulo-11-naive-bayes-infografia.png){download="chapter-11-naive-bayes-infographic.png"} |
| Google Colab notebook | Self-contained executable notebook. | [Open in Google Colab](https://colab.research.google.com/github/gilbertorodriguez59/libro-machine-learning-r/blob/main/colab/11-naive-bayes.ipynb){target="_blank"} | — |

::: {.content-visible when-format="html"}
<div class="video-responsive">
  <iframe src="https://www.youtube.com/embed/UbfJ3bLDwUI"
    title="Chapter 11. Naive Bayes"
    frameborder="0"
    allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share"
    referrerpolicy="strict-origin-when-cross-origin"
    allowfullscreen>
  </iframe>
</div>
:::

## Chapter summary

Naive Bayes provides a transparent probabilistic classifier based on Bayes' theorem and a simplifying conditional-independence assumption. Its simplicity is useful as a baseline, but smoothing, priors, sampling, and calibration all affect interpretation. Model-development decisions belong to training or validation data; the final test set is reserved for one independent evaluation.
