# Artificial Neural Networks

::: {.callout-note title="Connection with the theoretical volume"}
The mathematical treatment of linear algebra, calculus, gradient descent, backpropagation, and neural networks is developed in Chapters 2, 4, and 16 of *Mathematical Foundations of Machine Learning* [@rodriguez2026fundamentos].
:::

## Learning objectives

By the end of this chapter, the reader will be able to:

- explain the structure of an artificial neuron;
- interpret inputs, weights, bias, and activation functions;
- distinguish a perceptron from a multilayer network;
- describe forward propagation and backpropagation;
- match output activations and loss functions to the task;
- prepare data without leakage;
- implement a neuron and a perceptron without specialized packages;
- train a small network with `neuralnet`;
- select architecture and threshold with validation data;
- recognize the modern `keras3` interface for R;
- evaluate the selected network once on an untouched test set.

## Introduction

Artificial neural networks combine weighted linear transformations with nonlinear activation functions. Their flexibility makes them useful for classification, regression, images, signals, text, and time series.

That flexibility also increases the risk of overfitting. A network should therefore be judged by validation and test performance rather than by training fit alone.

## The artificial neuron

A neuron receives predictors

$$
x_1,x_2,\ldots,x_p
$$

with weights

$$
w_1,w_2,\ldots,w_p.
$$

It computes

$$
z=w_1x_1+w_2x_2+\cdots+w_px_p+b
$$

and then applies an activation function

$$
a=f(z).
$$

The parameter $b$ is the bias.

## Manual neuron example

Suppose

$$
x_1=0.8,\quad x_2=0.4,\quad
w_1=1.2,\quad w_2=-0.7,\quad b=0.1.
$$

Then

$$
z=(1.2)(0.8)+(-0.7)(0.4)+0.1=0.78.
$$

With the sigmoid activation,

$$
\sigma(z)=\frac{1}{1+e^{-z}},
$$

we obtain approximately $0.686$.

```{r manual-neuron}
sigmoid <- function(z) 1 / (1 + exp(-z))

x <- c(0.8, 0.4)
w <- c(1.2, -0.7)
b <- 0.1

z <- sum(x * w) + b
output <- sigmoid(z)

data.frame(z = z, output = output)
```

## Activation functions

### Sigmoid

$$
\sigma(z)=\frac{1}{1+e^{-z}}.
$$

It is commonly used for binary-classification outputs.

### Hyperbolic tangent

$$
\tanh(z)=\frac{e^z-e^{-z}}{e^z+e^{-z}}.
$$

### ReLU

$$
\operatorname{ReLU}(z)=\max(0,z).
$$

ReLU is widely used in hidden layers.

### Softmax

For $K$ classes,

$$
P(y=k)=\frac{e^{z_k}}{\sum_{j=1}^{K}e^{z_j}}.
$$

::: {.callout-tip title="Match the output to the task"}
A sigmoid output is common for binary classification, softmax for multiclass classification, and a linear output for ordinary regression.
:::

## Interactive laboratory: artificial neuron

::: {.content-visible when-format="html"}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 700
#| components: [viewer]

library(shiny)

sigmoid <- function(z) 1 / (1 + exp(-z))
relu <- function(z) pmax(0, z)

ui <- fluidPage(
  titlePanel("Interactive artificial neuron"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("x1", "Input x1", -5, 5, 0.8, 0.1),
      sliderInput("x2", "Input x2", -5, 5, 0.4, 0.1),
      sliderInput("w1", "Weight w1", -5, 5, 1.2, 0.1),
      sliderInput("w2", "Weight w2", -5, 5, -0.7, 0.1),
      sliderInput("b", "Bias b", -5, 5, 0.1, 0.1),
      selectInput(
        "activation", "Activation",
        choices = c("Sigmoid" = "sigmoid", "tanh" = "tanh",
                    "ReLU" = "relu", "Linear" = "linear")
      ),
      sliderInput("threshold", "Threshold", 0, 1, 0.5, 0.05)
    ),
    mainPanel(
      verbatimTextOutput("result"),
      plotOutput("plot", height = "390px")
    )
  )
)

server <- function(input, output, session) {
  z_now <- reactive(input$w1 * input$x1 + input$w2 * input$x2 + input$b)

  output_now <- reactive({
    switch(
      input$activation,
      sigmoid = sigmoid(z_now()),
      tanh = tanh(z_now()),
      relu = relu(z_now()),
      linear = z_now()
    )
  })

  output$result <- renderPrint({
    cat("z =", round(z_now(), 4), "\n")
    cat("output =", round(output_now(), 4), "\n")
    if (input$activation == "sigmoid") {
      cat("class =", ifelse(output_now() >= input$threshold, 1, 0))
    }
  })

  output$plot <- renderPlot({
    zz <- seq(-6, 6, length.out = 400)
    yy <- switch(
      input$activation,
      sigmoid = sigmoid(zz),
      tanh = tanh(zz),
      relu = relu(zz),
      linear = zz
    )
    plot(zz, yy, type = "l", lwd = 3,
         xlab = "z", ylab = "f(z)",
         main = paste("Activation:", input$activation))
    points(z_now(), output_now(), pch = 19, cex = 1.8)
    grid()
  })
}

shinyApp(ui, server)
```
:::

::: {.content-visible when-format="pdf"}
The interactive artificial-neuron laboratory is available in the web edition.
:::

## The perceptron

A perceptron uses a linear decision boundary:

$$
\widehat y=f(\mathbf w^T\mathbf x+b).
$$

A simple update rule is

$$
w_j^{new}=w_j^{old}+\eta(y-\widehat y)x_j.
$$

```{r perceptron-from-scratch}
train_perceptron <- function(X, y, eta = 0.1, epochs = 100) {
  X <- as.matrix(X)
  weights <- rep(0, ncol(X))
  bias <- 0

  for (epoch in seq_len(epochs)) {
    for (i in seq_len(nrow(X))) {
      z <- sum(X[i, ] * weights) + bias
      pred <- ifelse(z >= 0, 1, 0)
      error <- y[i] - pred
      weights <- weights + eta * error * X[i, ]
      bias <- bias + eta * error
    }
  }

  list(weights = weights, bias = bias)
}

X_and <- data.frame(x1 = c(0, 0, 1, 1), x2 = c(0, 1, 0, 1))
y_and <- c(0, 0, 0, 1)

and_model <- train_perceptron(X_and, y_and, epochs = 20)
and_model
```

A single perceptron cannot solve XOR because XOR is not linearly separable. Hidden layers allow a network to represent more complex boundaries.

## Multilayer networks

For one hidden layer,

$$
\mathbf h=f(\mathbf W^{(1)}\mathbf x+\mathbf b^{(1)}),
$$

followed by

$$
\widehat{\mathbf y}=g(\mathbf W^{(2)}\mathbf h+\mathbf b^{(2)}).
$$

## Loss functions

For regression,

$$
MSE=\frac{1}{n}\sum_{i=1}^{n}(y_i-\widehat y_i)^2.
$$

For binary classification,

$$
L=-\frac{1}{n}\sum_{i=1}^{n}
\left[y_i\log(\widehat p_i)+(1-y_i)\log(1-\widehat p_i)\right].
$$

## Gradient descent and backpropagation

A generic gradient-descent update is

$$
w^{new}=w^{old}-\eta\frac{\partial L}{\partial w}.
$$

Backpropagation applies the chain rule to propagate gradients from the output toward earlier layers. It is a training mechanism, not a separate predictive model.

## Data preparation

Neural networks require particular care with missing data, categorical encoding, scaling, class imbalance, and data partitioning.

::: {.callout-important title="Avoid data leakage"}
Scaling, imputation, feature selection, architecture selection, regularization, and threshold selection must be determined without using the final test set. The test set is reserved for one independent evaluation.
:::

## Binary example with `iris` and `neuralnet`

Training blocks are shown with `eval: false` to keep automated book rendering stable. They can be run directly in RStudio or in the chapter's Colab notebook.

### Create training, validation, and test sets

```{r iris-nn-prepare}
iris_bin <- subset(iris, Species != "setosa")
iris_bin$class <- ifelse(iris_bin$Species == "virginica", 1, 0)
iris_bin$Species <- NULL

stratified_sample <- function(y, proportion, seed) {
  set.seed(seed)
  unlist(lapply(
    split(seq_along(y), y),
    function(i) sample(i, floor(proportion * length(i)))
  ))
}

idx_tv <- stratified_sample(iris_bin$class, 0.80, 2026)
trainval_bin <- iris_bin[idx_tv, ]
test_bin <- iris_bin[-idx_tv, ]

idx_train <- stratified_sample(trainval_bin$class, 0.75, 2027)
train_bin <- trainval_bin[idx_train, ]
valid_bin <- trainval_bin[-idx_train, ]

predictors <- setdiff(names(train_bin), "class")

means_train <- sapply(train_bin[predictors], mean)
sds_train <- sapply(train_bin[predictors], sd)
sds_train[sds_train == 0] <- 1

train_bin[predictors] <- scale(train_bin[predictors], means_train, sds_train)
valid_bin[predictors] <- scale(valid_bin[predictors], means_train, sds_train)
```

### Metric function

```{r neural-network-metrics}
network_metrics <- function(actual, predicted) {
  tp <- sum(actual == 1 & predicted == 1)
  tn <- sum(actual == 0 & predicted == 0)
  fp <- sum(actual == 0 & predicted == 1)
  fn <- sum(actual == 1 & predicted == 0)
  safe_div <- function(a, b) if (b == 0) NA_real_ else a / b

  sensitivity <- safe_div(tp, tp + fn)
  specificity <- safe_div(tn, tn + fp)
  precision <- safe_div(tp, tp + fp)

  data.frame(
    accuracy = safe_div(tp + tn, tp + tn + fp + fn),
    sensitivity = sensitivity,
    specificity = specificity,
    balanced_accuracy = mean(c(sensitivity, specificity), na.rm = TRUE),
    precision = precision,
    f1 = if (is.na(precision) || is.na(sensitivity) ||
             precision + sensitivity == 0) NA_real_ else
      2 * precision * sensitivity / (precision + sensitivity)
  )
}
```

### Select architecture on validation data

```{r select-network-architecture}
#| eval: false

architectures <- list(c(3), c(5), c(5, 3))

architecture_results <- lapply(seq_along(architectures), function(i) {
  set.seed(3000 + i)

  model <- neuralnet::neuralnet(
    class ~ .,
    data = train_bin,
    hidden = architectures[[i]],
    linear.output = FALSE,
    lifesign = "none",
    stepmax = 1e6
  )

  prob <- as.numeric(
    neuralnet::compute(model, valid_bin[predictors])$net.result[, 1]
  )

  metrics <- network_metrics(valid_bin$class, ifelse(prob >= 0.5, 1, 0))

  cbind(
    architecture = paste(architectures[[i]], collapse = "-"),
    metrics
  )
})

do.call(rbind, architecture_results)
```

### Select the threshold on validation data

```{r select-network-threshold}
#| eval: false

validation_probability <- as.numeric(
  neuralnet::compute(validated_model, valid_bin[predictors])$net.result[, 1]
)

thresholds <- seq(0.20, 0.80, by = 0.05)

threshold_table <- do.call(rbind, lapply(thresholds, function(threshold) {
  cbind(
    threshold = threshold,
    network_metrics(
      valid_bin$class,
      ifelse(validation_probability >= threshold, 1, 0)
    )
  )
}))

threshold_table
```

::: {.callout-warning title="Do not tune the threshold on the test set"}
Trying many thresholds on `test_bin` and reporting the best one would make the test result optimistic. Architecture and threshold belong to validation; the test set is used once afterward.
:::

### Refit and evaluate once

After architecture and threshold selection, scaling parameters are recomputed from `trainval_bin`, the selected network is refitted on the development data, and `test_bin` is evaluated once.

```{r evaluate-final-network}
#| eval: false

means_final <- sapply(trainval_bin[predictors], mean)
sds_final <- sapply(trainval_bin[predictors], sd)
sds_final[sds_final == 0] <- 1

trainval_final <- trainval_bin
test_final <- test_bin

trainval_final[predictors] <- scale(
  trainval_final[predictors], means_final, sds_final
)
test_final[predictors] <- scale(
  test_final[predictors], means_final, sds_final
)

set.seed(2026)
final_network <- neuralnet::neuralnet(
  class ~ .,
  data = trainval_final,
  hidden = selected_architecture,
  linear.output = FALSE,
  lifesign = "none",
  stepmax = 1e6
)

test_probability <- as.numeric(
  neuralnet::compute(final_network, test_final[predictors])$net.result[, 1]
)
test_prediction <- ifelse(test_probability >= selected_threshold, 1, 0)

network_metrics(test_final$class, test_prediction)
```

## ATUS route

Categorical predictors must be converted to indicator variables before fitting a numerical neural network.

```{r atus-network-prepare}
library(readr)
library(dplyr)

atus <- read_csv("../datos/atus_ml_preparado.csv", show_col_types = FALSE)

set.seed(2026)
atus_nn <- atus |>
  sample_n(min(6000, nrow(atus))) |>
  mutate(
    y = ifelse(accidente_con_victimas == "Con víctimas", 1, 0),
    MES = factor(MES),
    DIASEMANA = factor(DIASEMANA),
    TIPACCID = factor(TIPACCID),
    CAUSAACCI = factor(CAUSAACCI),
    ID_HORA = as.numeric(ID_HORA)
  ) |>
  na.omit()

X_atus <- model.matrix(
  ~ ID_HORA + MES + DIASEMANA + TIPACCID + CAUSAACCI - 1,
  data = atus_nn
)

head(data.frame(y = atus_nn$y, X_atus)[, 1:8])
```

Continuous variables may be standardized using training parameters; 0/1 indicators can remain binary. The same transformation must be reproduced for validation and test sets.

## Interactive architecture visualizer

::: {.content-visible when-format="html"}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 720
#| components: [viewer]

library(shiny)

ui <- fluidPage(
  titlePanel("Neural-network architecture"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("p", "Inputs", 2, 8, 4, 1),
      sliderInput("h1", "Hidden layer 1", 1, 10, 5, 1),
      sliderInput("h2", "Hidden layer 2", 0, 8, 3, 1),
      sliderInput("out", "Outputs", 1, 4, 1, 1)
    ),
    mainPanel(
      plotOutput("network", height = "500px"),
      verbatimTextOutput("summary")
    )
  )
)

server <- function(input, output, session) {
  output$network <- renderPlot({
    layers <- c(input$p, input$h1)
    if (input$h2 > 0) layers <- c(layers, input$h2)
    layers <- c(layers, input$out)

    xx <- seq(0.08, 0.92, length.out = length(layers))
    positions <- lapply(seq_along(layers), function(j) {
      cbind(xx[j], seq(0.15, 0.85, length.out = layers[j]))
    })

    plot(0, 0, type = "n", xlim = c(0, 1), ylim = c(0, 1),
         axes = FALSE, xlab = "", ylab = "", main = "Topology")

    for (j in seq_len(length(positions) - 1)) {
      for (a in seq_len(nrow(positions[[j]]))) {
        for (b in seq_len(nrow(positions[[j + 1]]))) {
          segments(positions[[j]][a, 1], positions[[j]][a, 2],
                   positions[[j + 1]][b, 1], positions[[j + 1]][b, 2],
                   col = "gray70")
        }
      }
    }

    for (points_layer in positions) {
      points(points_layer[, 1], points_layer[, 2],
             pch = 21, bg = "white", cex = 2.3)
    }
  })

  output$summary <- renderPrint({
    hidden <- if (input$h2 > 0) c(input$h1, input$h2) else input$h1
    cat("hidden = ", paste(hidden, collapse = ", "), "\n", sep = "")
    cat("outputs = ", input$out, "\n", sep = "")
  })
}

shinyApp(ui, server)
```
:::

## Overfitting and regularization

Common controls include smaller architectures, L1/L2 regularization, dropout, early stopping, more data when appropriate, and repeated validation or cross-validation when computationally feasible.

## Modern implementation with `keras3`

The current Keras interface for R is provided by the `keras3` package. The following block is not executed automatically because it requires a configured backend.

```{r keras3-example}
#| eval: false

library(keras3)

x_train_keras <- as.matrix(trainval_final[predictors])
y_train_keras <- as.integer(trainval_final$class)

keras_model <- keras_model_sequential(input_shape = ncol(x_train_keras)) |>
  layer_dense(units = 16, activation = "relu") |>
  layer_dropout(rate = 0.20) |>
  layer_dense(units = 8, activation = "relu") |>
  layer_dense(units = 1, activation = "sigmoid")

keras_model |>
  compile(
    optimizer = "adam",
    loss = "binary_crossentropy",
    metrics = "accuracy"
  )

history <- keras_model |>
  fit(
    x_train_keras,
    y_train_keras,
    epochs = 100,
    batch_size = 16,
    validation_split = 0.20,
    verbose = 0
  )
```

## COVID-19 educational route

The processed period is selected through the same reproducible procedure used in the data-preparation chapter.

```{r covid-network-prepare}
comparison <- read_csv(
  "../datos/covid19/comparacion_estructural_2020_2025.csv",
  show_col_types = FALSE
)

selected_year <- comparison |>
  filter(is_usable_for_book, !is.na(prepared_ml_file)) |>
  arrange(desc(common_columns), desc(rows)) |>
  slice(1) |>
  pull(year)

covid_file <- file.path(
  "../datos/covid19/procesados",
  paste0("covid19_mexico_", selected_year, "_ml_preparado.csv.gz")
)

covid_nn <- read_csv(covid_file, show_col_types = FALSE) |>
  select(
    MURIO, EDAD, NEUMONIA, DIABETES, HIPERTENSION,
    OBESIDAD, RENAL_CRONICA, NUM_COMORBILIDADES
  ) |>
  tidyr::drop_na()

set.seed(2026)
covid_nn <- covid_nn |>
  sample_n(min(8000, nrow(covid_nn)))
```

### Development and test split

```{r covid-network-split}
idx_covid_tv <- stratified_sample(covid_nn$MURIO, 0.80, 2026)
covid_trainval <- covid_nn[idx_covid_tv, ]
covid_test <- covid_nn[-idx_covid_tv, ]

idx_covid_train <- stratified_sample(covid_trainval$MURIO, 0.75, 2027)
covid_train <- covid_trainval[idx_covid_train, ]
covid_valid <- covid_trainval[-idx_covid_train, ]

continuous <- c("EDAD", "NUM_COMORBILIDADES")
means_covid <- sapply(covid_train[continuous], mean)
sds_covid <- sapply(covid_train[continuous], sd)
sds_covid[sds_covid == 0] <- 1

covid_train[continuous] <- scale(covid_train[continuous], means_covid, sds_covid)
covid_valid[continuous] <- scale(covid_valid[continuous], means_covid, sds_covid)
```

Binary variables remain 0/1; only continuous variables are standardized. Architecture and threshold belong to `covid_valid`, after which the selected network is refitted on `covid_trainval` and evaluated once on `covid_test`.

::: {.callout-important title="Probabilities from an educational sample"}
If the sample deliberately changes the proportion of recorded deaths, sigmoid outputs must not automatically be interpreted as population risk. Natural prevalence or appropriate weighting and calibration are needed for population-oriented probability interpretation.
:::

::: {.callout-warning title="Educational use only"}
This example is not a diagnostic, prognostic, or clinical decision tool.
:::

## Comparison with simpler models

| Aspect | Neural network | Logistic regression | Decision tree | Random Forest |
|---|---|---|---|---|
| Nonlinearity | High | Limited without transformations | High | High |
| Interpretability | Low | High | High | Medium |
| Scaling | Important | Recommended | Usually unnecessary | Usually unnecessary |
| Data needs | Often larger | Smaller | Moderate | Moderate |
| Computational cost | Medium-high | Low | Low | Medium-high |

A more complex network is justified only when it provides a meaningful and stable improvement over simpler alternatives.

## Common mistakes

1. Scaling before splitting the data.
2. Selecting architecture on the test set.
3. Selecting the threshold on the test set.
4. Training too many epochs without validation.
5. Reporting only accuracy.
6. Ignoring class imbalance.
7. Failing to record seeds and software versions.
8. Treating sigmoid output as certainty.
9. Failing to compare against a baseline.

## Supplementary materials for this chapter

| Resource | Purpose | Open or play | Download |
|---|---|---|---|
| Explanatory video | Audiovisual introduction to artificial neural networks. | [Watch on YouTube](https://www.youtube.com/watch?v=gk3bt0h3WUU){target="_blank"} | — |
| PDF presentation | Slides for study or classroom presentation. | [View PDF](../recursos/capitulo-12/capitulo-12-redes-neuronales-presentacion.pdf){target="_blank"} | [Download PDF](../recursos/capitulo-12/capitulo-12-redes-neuronales-presentacion.pdf){download="chapter-12-neural-networks-presentation.pdf"} |
| Editable presentation | PowerPoint file for classroom use or adaptation. | [Open PPTX](../recursos/capitulo-12/capitulo-12-redes-neuronales-presentacion.pptx) | [Download PPTX](../recursos/capitulo-12/capitulo-12-redes-neuronales-presentacion.pptx){download="chapter-12-neural-networks-presentation.pptx"} |
| Infographic | Visual summary of the main ideas and architecture. | [View infographic](../recursos/capitulo-12/capitulo-12-redes-neuronales-infografia.png){target="_blank"} | [Download PNG](../recursos/capitulo-12/capitulo-12-redes-neuronales-infografia.png){download="chapter-12-neural-networks-infographic.png"} |
| Google Colab notebook | Self-contained executable notebook. | [Open in Google Colab](https://colab.research.google.com/github/gilbertorodriguez59/libro-machine-learning-r/blob/main/colab/12-redes-neuronales.ipynb){target="_blank"} | — |

::: {.content-visible when-format="html"}
<div class="video-responsive">
  <iframe src="https://www.youtube.com/embed/gk3bt0h3WUU"
    title="Chapter 12. Artificial Neural Networks"
    frameborder="0"
    allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share"
    referrerpolicy="strict-origin-when-cross-origin"
    allowfullscreen>
  </iframe>
</div>
:::

## Chapter summary

Neural networks combine weighted transformations and nonlinear activations to represent complex relationships. Their flexibility does not remove the need for rigorous validation: preprocessing, architecture, regularization, and threshold selection belong to model development; the final test set must remain independent. A network adds value only when it generalizes and offers a sufficient improvement over simpler alternatives.
