# Clustering with k-means

::: {.callout-note title="Connection with the theoretical volume"}
The mathematical treatment of distances, centroids, objective functions, and k-means clustering is developed in Chapters 2 and 18 of *Mathematical Foundations of Machine Learning* [@rodriguez2026fundamentos].
:::

## Learning objectives

By the end of this chapter, the reader will be able to:

- distinguish supervised from unsupervised learning;
- explain the k-means algorithm and its objective function;
- interpret centroids and assignments;
- understand the effect of scaling and initialization;
- compare values of $k$ using elbow and silhouette diagnostics;
- inspect stability across random initializations;
- interpret cluster profiles without treating clusters as true classes;
- apply k-means to real ATUS profiles and an educational COVID-19 sample.

## From supervised to unsupervised learning

In supervised learning, a known response guides model fitting. In clustering, no class label defines the correct grouping. The algorithm discovers structure using the selected variables and a definition of similarity.

::: {.callout-important title="A cluster is not an external truth"}
The result depends on the variables, scaling, distance, value of $k$, and sample. A cluster should not automatically be treated as a natural or causal real-world category.
:::

## Core idea

k-means partitions observations into $k$ groups. Each group is represented by a centroid. The algorithm alternates between:

1. assigning each observation to the nearest centroid;
2. recomputing centroids from the current assignments;
3. repeating until the solution stabilizes.

## Objective function

k-means minimizes the within-cluster sum of squares:

$$
WSS=\sum_{j=1}^{k}\sum_{\mathbf{x}_i\in C_j}
\|\mathbf{x}_i-\boldsymbol{\mu}_j\|^2,
$$

where $C_j$ is cluster $j$ and $\boldsymbol{\mu}_j$ is its centroid.

## Basic R example

```{r kmeans-basic}
data <- data.frame(
  x = c(1, 1.5, 3, 5, 3.5, 4.5),
  y = c(1, 2, 4, 7, 5, 5)
)

set.seed(123)
km <- kmeans(data, centers = 2, nstart = 25)
km
```

```{r kmeans-basic-plot, fig.width=7, fig.height=4.8}
data$cluster <- factor(km$cluster)
plot(
  data$x, data$y,
  col = data$cluster,
  pch = 19,
  xlab = "x",
  ylab = "y",
  main = "k-means clustering"
)
points(km$centers, pch = 8, cex = 2.2, lwd = 2)
```

## Why multiple random starts matter

k-means can converge to different local solutions. `nstart = 25` runs several initializations and retains the solution with the smallest within-cluster variation.

::: {.callout-tip title="Initialization and stability"}
A large `nstart` reduces dependence on one initialization, but it does not replace stability checks across samples, variables, or preprocessing choices.
:::

## Scaling

Because k-means is distance-based, scale defines the geometry of the problem. Continuous variables are often standardized:

$$
z=\frac{x-\bar{x}}{s}.
$$

```{r kmeans-iris-scale}
X_iris <- scale(iris[, 1:4])
head(X_iris)
```

::: {.callout-important title="Standardizing binary variables is a weighting decision"}
A 0/1 variable does not have to be standardized automatically. Standardization can give a rare category substantial relative influence. With mixed numeric and binary data, the intended weighting should be explicit and documented.
:::

## `iris` example

```{r kmeans-iris}
set.seed(2026)
km_iris <- kmeans(X_iris, centers = 3, nstart = 50)

table(
  Cluster = km_iris$cluster,
  Species = iris$Species
)
```

Species was not used to form the clusters. It is shown afterward only to compare unsupervised structure with a known label.

## Centroids

```{r kmeans-iris-centroids}
km_iris$centers
```

With standardized variables, positive centroid values indicate averages above the overall mean and negative values indicate averages below it.

## Choosing $k$: elbow method

```{r kmeans-elbow, fig.width=7, fig.height=4.5}
k_values <- 1:10
wss <- sapply(k_values, function(k_now) {
  set.seed(2026)
  kmeans(X_iris, centers = k_now, nstart = 30)$tot.withinss
})

plot(
  k_values, wss,
  type = "b", pch = 19,
  xlab = "Number of clusters k",
  ylab = "Within-cluster sum of squares",
  main = "Elbow method"
)
```

The elbow is the region where additional clusters begin to provide much smaller reductions in WSS. It is a diagnostic, not an automatic rule.

## Silhouette coefficient

Silhouette compares cohesion within the assigned cluster with separation from the nearest competing cluster. Values close to 1 indicate clear membership, values near 0 indicate boundary observations, and negative values suggest possible misassignment.

```{r kmeans-iris-silhouette}
d_iris <- dist(X_iris)

sil_iris <- sapply(2:6, function(k_now) {
  set.seed(2026)
  model <- kmeans(X_iris, centers = k_now, nstart = 30)
  sil <- cluster::silhouette(model$cluster, d_iris)
  mean(sil[, "sil_width"])
})

data.frame(k = 2:6, mean_silhouette = sil_iris)
```

::: {.callout-note title="There is no universally correct k"}
A reasonable choice combines elbow, silhouette, stability, interpretability, and domain knowledge. A tiny numerical advantage does not necessarily justify a much more complicated segmentation.
:::

## Stability across initializations

```{r kmeans-iris-stability}
seed_results <- do.call(rbind, lapply(1:20, function(s) {
  set.seed(s)
  model <- kmeans(X_iris, centers = 3, nstart = 10)
  data.frame(
    seed = s,
    wss = model$tot.withinss,
    between_ratio = model$betweenss / model$totss
  )
}))

summary(seed_results[, c("wss", "between_ratio")])
```

Large changes across seeds despite multiple starts indicate that the solution deserves additional caution.

## Interactive k-means laboratory

::: {.content-visible when-format="html"}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 690
#| components: [viewer]

library(shiny)

ui <- fluidPage(
  titlePanel("Interactive k-means laboratory"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("k", "Number of clusters", 2, 6, 3, 1),
      sliderInput("n", "Number of observations", 30, 200, 90, 10),
      sliderInput("spread", "Dispersion", 0.2, 2.0, 0.7, 0.1),
      numericInput("seed", "Seed", 123, 1),
      actionButton("update", "Generate and cluster")
    ),
    mainPanel(
      plotOutput("plot", height = "450px"),
      tableOutput("summary")
    )
  )
)

server <- function(input, output, session) {
  data_sim <- eventReactive(input$update, {
    set.seed(input$seed)
    centers <- matrix(
      c(-3,-2, 0,3, 3,-1, -2,4, 4,4, 0,-4),
      ncol = 2, byrow = TRUE
    )
    g <- sample(seq_len(input$k), input$n, replace = TRUE)
    data.frame(
      x = centers[g, 1] + rnorm(input$n, sd = input$spread),
      y = centers[g, 2] + rnorm(input$n, sd = input$spread)
    )
  }, ignoreNULL = FALSE)

  fit <- reactive({
    set.seed(input$seed)
    kmeans(data_sim(), centers = input$k, nstart = 25)
  })

  output$plot <- renderPlot({
    d <- data_sim(); model <- fit()
    plot(d$x, d$y, col = model$cluster, pch = 19,
         xlab = "Variable 1", ylab = "Variable 2",
         main = paste("k-means with k =", input$k))
    points(model$centers, pch = 8, cex = 2.2, lwd = 3)
    grid()
  })

  output$summary <- renderTable({
    model <- fit()
    data.frame(
      Cluster = seq_len(input$k),
      Size = as.integer(model$size),
      Center_X = round(model$centers[, 1], 2),
      Center_Y = round(model$centers[, 2], 2)
    )
  })
}

shinyApp(ui, server)
```
:::

::: {.content-visible when-format="pdf"}
The interactive k-means laboratory is available in the web edition.
:::

## Applied case A: monthly ATUS profiles

Rather than clustering raw categorical accident records directly, we first construct **numeric monthly profiles** from the prepared ATUS dataset.

```{r atus-kmeans-prepare}
library(readr)
library(dplyr)

atus <- read_csv("../datos/atus_ml_preparado.csv", show_col_types = FALSE)

atus_monthly <- atus |>
  mutate(
    MES = as.numeric(MES),
    ID_HORA = as.numeric(ID_HORA),
    weekend = DIASEMANA %in% c(
      "sábado", "sabado", "domingo",
      "Sábado", "Sabado", "Domingo"
    ),
    casualties = accidente_con_victimas == "Con víctimas"
  ) |>
  group_by(MES) |>
  summarise(
    accidents = n(),
    mean_hour = mean(ID_HORA, na.rm = TRUE),
    weekend_proportion = mean(weekend, na.rm = TRUE),
    casualty_proportion = mean(casualties, na.rm = TRUE),
    accident_types = n_distinct(TIPACCID),
    .groups = "drop"
  ) |>
  filter(if_all(-MES, is.finite))

atus_monthly
```

Each row is now a month rather than an individual accident.

```{r atus-kmeans-scale}
atus_vars <- c(
  "accidents", "mean_hour", "weekend_proportion",
  "casualty_proportion", "accident_types"
)

X_atus <- scale(atus_monthly[atus_vars])
```

### Choose among a small range of k

With only 12 months, a large number of clusters would not be meaningful.

```{r atus-kmeans-select-k}
d_atus <- dist(X_atus)

sil_atus <- sapply(2:4, function(k_now) {
  set.seed(2026)
  model <- kmeans(X_atus, centers = k_now, nstart = 50)
  mean(cluster::silhouette(model$cluster, d_atus)[, "sil_width"])
})

atus_k_table <- data.frame(k = 2:4, mean_silhouette = sil_atus)
atus_k_table

k_atus <- atus_k_table$k[which.max(atus_k_table$mean_silhouette)]
```

```{r atus-kmeans-model}
set.seed(2026)
km_atus <- kmeans(X_atus, centers = k_atus, nstart = 100)
atus_monthly$cluster <- factor(km_atus$cluster)

atus_monthly |>
  arrange(cluster, MES)
```

```{r atus-kmeans-profiles}
atus_monthly |>
  group_by(cluster) |>
  summarise(
    months = paste(MES, collapse = ", "),
    mean_accidents = mean(accidents),
    mean_hour = mean(mean_hour),
    mean_weekend = mean(weekend_proportion),
    mean_casualties = mean(casualty_proportion),
    .groups = "drop"
  )
```

::: {.callout-warning title="Only 12 aggregated observations"}
This is a teaching example for constructing real ATUS profiles. It does not support strong claims about seasonal structure; the clusters describe these 12 monthly profiles in the prepared dataset.
:::

## Applied case B: COVID-19 profiles

The outcome `MURIO` is excluded from cluster formation and is used only afterward as an external descriptive variable.

```{r covid-kmeans-prepare}
comparison <- read_csv(
  "../datos/covid19/comparacion_estructural_2020_2025.csv",
  show_col_types = FALSE
)

selected_year <- comparison |>
  filter(is_usable_for_book, !is.na(prepared_ml_file)) |>
  arrange(desc(common_columns), desc(rows)) |>
  slice(1) |>
  pull(year)

covid_file <- file.path(
  "../datos/covid19/procesados",
  paste0("covid19_mexico_", selected_year, "_ml_preparado.csv.gz")
)

covid_km <- read_csv(covid_file, show_col_types = FALSE) |>
  select(
    MURIO, EDAD, NEUMONIA, DIABETES, HIPERTENSION,
    OBESIDAD, RENAL_CRONICA
  ) |>
  tidyr::drop_na()

set.seed(2026)
covid_km <- covid_km |>
  sample_n(min(12000, nrow(covid_km)))
```

### Build the clustering matrix

Only age is standardized in this edition. Binary comorbidity variables remain 0/1. This avoids automatically giving a rare binary variable a large weight simply because its variance is small.

```{r covid-kmeans-matrix}
X_covid <- covid_km |>
  transmute(
    EDAD = as.numeric(scale(EDAD)),
    NEUMONIA,
    DIABETES,
    HIPERTENSION,
    OBESIDAD,
    RENAL_CRONICA
  ) |>
  as.matrix()
```

::: {.callout-note title="Weighting is part of the model"}
Standardizing every variable or assigning explicit weights would define a different geometry and could produce different clusters. The choice should be documented rather than treated as a hidden preprocessing default.
:::

### Diagnose k on a reproducible subsample

```{r covid-kmeans-select-k}
set.seed(2026)
n_sil <- min(2500, nrow(X_covid))
idx_sil <- sample(seq_len(nrow(X_covid)), n_sil)
X_sil <- X_covid[idx_sil, , drop = FALSE]
d_sil <- dist(X_sil)

k_candidates <- 2:6
sil_covid <- sapply(k_candidates, function(k_now) {
  set.seed(2026)
  model <- kmeans(X_sil, centers = k_now, nstart = 40)
  mean(cluster::silhouette(model$cluster, d_sil)[, "sil_width"])
})

covid_k_table <- data.frame(
  k = k_candidates,
  mean_silhouette = sil_covid
)
covid_k_table

k_covid <- covid_k_table$k[which.max(covid_k_table$mean_silhouette)]
k_covid
```

### Elbow method

```{r covid-kmeans-elbow, fig.width=7, fig.height=4.5}
wss_covid <- sapply(1:8, function(k_now) {
  set.seed(2026)
  kmeans(X_covid, centers = k_now, nstart = 30)$tot.withinss
})

plot(
  1:8, wss_covid,
  type = "b", pch = 19,
  xlab = "Number of clusters k",
  ylab = "Within-cluster sum of squares",
  main = "COVID-19: elbow method"
)
```

Silhouette provides a quantitative guide and the elbow offers a complementary perspective. If they disagree strongly, the ambiguity should be reported rather than hidden.

### Fit the selected solution

```{r covid-kmeans-model}
set.seed(2026)
km_covid <- kmeans(
  X_covid,
  centers = k_covid,
  nstart = 100,
  iter.max = 100
)

covid_km$cluster <- factor(km_covid$cluster)
table(covid_km$cluster)
```

### Stability across seeds

```{r covid-kmeans-stability}
covid_stability <- do.call(rbind, lapply(1:20, function(s) {
  set.seed(s)
  model <- kmeans(X_covid, centers = k_covid, nstart = 20)
  data.frame(
    seed = s,
    wss = model$tot.withinss,
    between_ratio = model$betweenss / model$totss
  )
}))

summary(covid_stability[, c("wss", "between_ratio")])
```

This checks initialization stability. A stronger analysis would also repeat the procedure on different subsamples.

### Profiles in original units

```{r covid-kmeans-profiles}
covid_profiles <- covid_km |>
  group_by(cluster) |>
  summarise(
    n = n(),
    mean_age = mean(EDAD),
    pneumonia_pct = mean(NEUMONIA) * 100,
    diabetes_pct = mean(DIABETES) * 100,
    hypertension_pct = mean(HIPERTENSION) * 100,
    obesity_pct = mean(OBESIDAD) * 100,
    chronic_kidney_pct = mean(RENAL_CRONICA) * 100,
    .groups = "drop"
  )

covid_profiles
```

### Describe `MURIO` after clustering

```{r covid-kmeans-outcome}
covid_km |>
  group_by(cluster) |>
  summarise(
    records = n(),
    sample_death_percentage = mean(MURIO == 1) * 100,
    .groups = "drop"
  )
```

::: {.callout-warning title="External variable, not a target"}
Differences in `MURIO` across clusters do not turn k-means into a classifier and do not establish causality. The variable did not participate in cluster formation.
:::

### PCA visualization

PCA is used only to create a two-dimensional view. Clustering remains defined by `X_covid`.

```{r covid-kmeans-pca, fig.width=7, fig.height=5}
pca_covid <- prcomp(X_covid, center = TRUE, scale. = TRUE)

covid_plot <- data.frame(
  PC1 = pca_covid$x[, 1],
  PC2 = pca_covid$x[, 2],
  cluster = covid_km$cluster
)

ggplot2::ggplot(
  covid_plot,
  ggplot2::aes(x = PC1, y = PC2, color = cluster)
) +
  ggplot2::geom_point(alpha = 0.35, size = 1.2) +
  ggplot2::labs(
    title = "COVID-19 profiles obtained with k-means",
    subtitle = "PCA is used only for visualization",
    x = "Principal component 1",
    y = "Principal component 2",
    color = "Cluster"
  ) +
  ggplot2::theme_minimal()
```

::: {.callout-warning title="Educational use only"}
COVID-19 clusters are statistical profiles within the selected sample. They are not diagnoses, clinical categories, or medical risk levels.
:::

## Strengths and limitations

### Strengths

- simple and intuitive;
- computationally efficient in many datasets;
- easy to summarize through centroids;
- useful for exploration and segmentation.

### Limitations

- requires choosing $k$;
- depends on scale and initialization;
- is sensitive to outliers;
- tends to favor compact, roughly spherical groups;
- is not naturally designed for purely categorical variables;
- can produce apparently clear clusters even when the underlying structure is weak.

## Common mistakes

1. Failing to document scaling.
2. Choosing $k$ only because it matches an expectation.
3. Using a single initialization.
4. Treating labels 1, 2, and 3 as ordered categories.
5. Including identifiers.
6. Calling clusters "true classes."
7. Describing clusters without inspecting centroids and distributions.
8. Using an outcome variable to form clusters and later presenting it as external validation.

## Recommended workflow

1. Define what observations should be grouped.
2. Select substantively meaningful variables.
3. Handle missing values and outliers.
4. Decide and document scaling or weighting.
5. Explore several values of $k$.
6. Inspect silhouette and elbow diagnostics.
7. Examine stability.
8. Fit with multiple random starts.
9. Interpret centroids in understandable units.
10. Validate whether the profiles are useful with domain knowledge.

## Supplementary materials for this chapter

| Resource | Purpose | Open or play | Download |
|---|---|---|---|
| Explanatory video | Audiovisual explanation of k-means clustering. | [Watch on YouTube](https://www.youtube.com/watch?v=mZgUlnSxrgA){target="_blank"} | — |
| PDF presentation | Slides for study or classroom presentation. | [View PDF](../recursos/capitulo-13/capitulo-13-kmeans-presentacion.pdf){target="_blank"} | [Download PDF](../recursos/capitulo-13/capitulo-13-kmeans-presentacion.pdf){download="chapter-13-kmeans-presentation.pdf"} |
| Editable presentation | PowerPoint file for classroom use or adaptation. | [Open PPTX](../recursos/capitulo-13/capitulo-13-kmeans-presentacion.pptx) | [Download PPTX](../recursos/capitulo-13/capitulo-13-kmeans-presentacion.pptx){download="chapter-13-kmeans-presentation.pptx"} |
| Infographic | Visual summary of the algorithm and modeling choices. | [View infographic](../recursos/capitulo-13/capitulo-13-kmeans-infografia.png){target="_blank"} | [Download PNG](../recursos/capitulo-13/capitulo-13-kmeans-infografia.png){download="chapter-13-kmeans-infographic.png"} |
| Google Colab notebook | Self-contained executable notebook. | [Open in Google Colab](https://colab.research.google.com/github/gilbertorodriguez59/libro-machine-learning-r/blob/main/colab/13-kmeans.ipynb){target="_blank"} | — |

::: {.content-visible when-format="html"}
<div class="video-responsive">
  <iframe src="https://www.youtube.com/embed/mZgUlnSxrgA"
    title="Chapter 13. k-means Clustering"
    frameborder="0"
    allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share"
    referrerpolicy="strict-origin-when-cross-origin"
    allowfullscreen>
  </iframe>
</div>
:::

## Chapter summary

k-means discovers structure through distances and centroids. A sound analysis does not end with `kmeans()`: variables, scaling, and the number of clusters must be justified; stability should be examined; and profiles should be interpreted cautiously. This chapter closes the book's main algorithmic route and prepares the ground for an integrated comparison of the methods studied.
